---
title: "Part 0: R for beginner"
output: html_document
author: "Marc Garel, MIO"
editor_options: 
  chunk_output_type: inline
---

## How to install R?
### Useful link

To install R, you can go to [https://cran.r-project.org/](https://cran.r-project.org/), 
and download the install file for your favorite operating system,
click on the .exe, .dmg, .pkg, .deb, respectively for Windows, MacOS and Linux-debian.

And click on follow... until to reach successful installation
Then download and install the IDE Rstudio
[https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/).

*Everything is free to download*

### First commands

**Where am I?**

To get the current working directory

```{r}
getwd()
```

**To change my working directory from the console**

```{r eval=FALSE}
setwd("/path/to/my/fancy/project/")
```

In Rstudio, we can change the working directory by navigating in folder in **File** panel
and clicking on menu **More**

If you are using a Rproject, you don't need to change your working directory.

**How can I find documentation about function?**

The command `help()` is the 911

```{r eval=FALSE}
# exemple with function read.table
help("read.table")
#other exemple of 911
example("plot")
??plot()
```


## How to use a package?

### Installation

You can install new packages by clicking directly in Rstudio
or by command line (the best way for me)

```{r eval=FALSE}
install.packages("your_package") # for packages on CRAN mirror
```

For packages from Bioconductor (specifically for bioinformatic): 

```{r eval=FALSE}
BiocManager::install("your_package")
```

or from github (using the package devtools)

```{r eval=FALSE}
devtools::install_github("your_package")
```

### Using functions from a package

To use functions from a specific package you can either load the entire package:

```{r eval=FALSE}
library("your package1")
library("your package2")
library("your package3")
```

or call the function this way:

```{r eval=FALSE}
your_package::yourfunction()
```

**<span style="color: red;">On your R session using cloud IFB all the necessary packages are installed</span>**

## Expression and affectation
### Expression

An expression is directly evaluated and the result is displayed on terminal
Example : 

```{r}
2 + 3
sqrt(25)
```


### Affectation in an object
An assignment is an expression stored in object or variable.
In this example expression, constant, array, matrix, data frame, list
Example :

```{r}
a <- 2 + 3
b <- "madame"
```

**<span style="color: red;"> What happens when you execute this cell?</span>**

```{r}
a
b
```

```{r}
a <- 10
b <- "5"
```

```{r eval=FALSE}
somme <- a + b
```

**<span style="color: red;"> Why do we have an error ? I need to know the type of the object</span>**
```{r}
str(a)
str(b)
```
**<span style="color: red;"> What kind of information do we get?</span>**

We can also perform mathematical operations on numerical objects.

```{r}
log(a) # to obtain the logarithm of a
sqrt(a)# to obtain the square root of a
```

```{r}
# we can compare it
log(a) > sqrt(a)
```

### Type of objects

#### Vectors

Vectors are objects composed by values with the same type or (i.e, numeric, characters...)

```{r}
V1 <- c(2, 6, 9) # numeric vector
V2 <- c("monday", "Tuesday", "Wednesday") # character vector
V3 <- rep(6, 3) # repetition of the same value
V4 <- seq(1, 3, 0.1)# sequence of number
V5 <- 1:100
V5
```


To know the number of value inside a vector

```{r}
length(V1)
```

**What do you think about V4?**

Filter a vector according to criteria

```{r}

# Example : with list of value (a vector)

x <- c(1, 3, 5, 3, 2, 1, 4, 6, 4, 7, 5, 4, 3)

# get element from 2 to 6

x[2:6]

# get elements 3 et 5 from x.

x[c(3, 5)]

# get value more than 20.

x[x > 5]

# get value of  x where x is equal to 21.

x[x == 5]

# return elements form x the the value different from 5

x[x != 5]
```

Filter a vector according to several criteria

```{r}
# 3 lists : ages, sexes et poids

age <- c(20, 30, 40,
         15, 22, 24,
         36, 38)

sexe <- c("F", "M", "F",
          "M", "F", "M",
          "F", "M")

poids <- c(75, 76, 73,
           72, 64, 76,
           73, 72)

# get value from age greater than 20 and less than 30.

age[age > 20 & age < 30]

# Recovering "poids" for those who are older than 25 and female

poids[age > 25 & sexe == "F"]

#Retrieve age values below 20 or above 30.

age[age < 20 | age > 30]
```

*Exo1*

Considering the vector `a` such as 
`a <- c("lannister", "targaryen", "baratheon", "starck", "greyjoy")`

1. What is the length of the vector?

2. Try doing `a[1:3]`. What do you get?

3. Create a new vector b containing only lannister and starck.

4. Try doing `a[-1]`. What do you get?

5. Sort by alphabetical order using `sort()`

*Exo2*

1. Create a vector `a` containing all integers from 1 to 100.

2. Add the values 200, 201, 202 to the vector `a`.

3. Create a vector `b` containing all even integers from 2 to 100 using `seq()`


#### Data frames

Data frames are objects composed by vector where the value
are of different modes (i.e, numeric, characters...)

##### Data frame examples

Load a data frame

```{r}
data(iris)
```

Visualise the data frame in a table

```{r eval=FALSE}
View(iris)
```

Display its internal structure

```{r}
str(iris)
```

**What can we notice?** 

##### How to build your own data frame

```{r}
date <- c("1_monday", "2_Tuesday", "3_Wednesday",
          "4_Thursday", "5_Friday", "6_Sturday",
          "7_Sunday")

is.character(date)

# temperature in deg Celsius
temperature <- c(24, 27, 25,
               22, 30, 21,
               28)

is.numeric(temperature)

# rain in mm
rain <- c(1, 0, 0,
          5, 2, 0,
          0)

is.numeric(rain)

# make data.frame
df <- data.frame(date, temperature, rain)
str(df)

#To select a column or vector
df$temperature
df[, 2]

# here we use list() instead of c()
# because there is multiple class in inside row

day <- list("8_monday", 29, 1)
new_def <- rbind(df, day)# add row to a data frame
new_def
```

## Functions
### Definition
Function are a compilation of command line with different instructions 
inside one object to simplify code. A function is composed by
 <span style="color: steelblue;">**arguments**</span> and <span style="color: steelblue;">**options**</span>.

function(argument1, argument2, option1, ... ,option10)


### Usual functions for data frame

```{r, eval=FALSE}
head() # to know first line of your data frame
class()# return the class of the object. Ex : data.frame, matrix, list ....
str()# return the structure of the object. Ex : numeric, factor, character....
names()# to get or set the names of an object
sum() # for addition
min() # return the minimum of the vector
max() # return the minimum of the vector
row.names() # attribute names for lines of the data frame
colnames() # attribute names for column of the data frame
apply() # Returns a vector or array or list of values obtained by applying a function to margins of an array or matrix.
```

**Example function row.names, class and str**

```{r}
df2 <- data.frame(x = c(TRUE, FALSE, NA, NA), y = c(12, 34, 56, 78))
df2
row.names(df2) <- paste("row", 1 : 4, sep = "_")
df2 # what do you see
class(df2)
str(df2)
```

**Example function apply**

```{r}
head(df)
class(df)
str(df)
# return mean for the numerical column of the data.frame. apply(data,margin,fun). 
# For margin parameter the value 1 return mean for each row, 
# for margin=2 return mean for each selected column.
apply(df[, 2:3], 2, mean)
```
*Exo 3*

**From data set iris in package **datasets** Load package datasets and load data set iris using `data()`

1.  Give the class Sepal.Width and Species vectors
2.  What is the minimum / maximum / average sepal length of these irises?
3.  What are the values of the first 10 irises?
4.  Calculate standard deviation for every numeric vector (function : `sd()`)
5.  Calculate mean for every numeric vector
6.  Create a data frame with mean and sd as line and give a name for each line
7.  An error of 0.5cm was made in the measurement of the length of the sepal of the 41st iris: add 0.5cm to this measurement

### Use dplyr to select, filter a data frame

`dplyr` is part of the library set named `tidyverse`
(contraction of "tidy" and "universe", it's a tidy universe). 
`tidyverse` packages are designed to work together and thus follow
the same code logic and a common grammar.

The pipe, `%>%`, is one of the useful elements of the tidyverse. 
It allows to structure sequences of operations by minimizing 
the creation of intermediate objects and by facilitating 
the addition of a step anywhere in this sequence. Note that from
R 4.1 you can use a new pipe, `|>` without the need of loading any
library.

The most commonly used tidyverse packages are loaded in your session:

- [`ggplot2`](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf)
- [`dplyr`](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf)
- [`tidyr`](https://github.com/rstudio/cheatsheets/blob/main/tidyr.pdf)
- [`readr`](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-import.pdf)
- [`tibble`](https://r4ds.had.co.nz/tibbles.html)
- [`stringr`](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf)



```{r}
tidyverse::tidyverse_packages()
```
##### Filter and select variable in data frame

```{r}
library(tidyverse)
data("iris")
#To select a variable with tidyverse
dplyr::select(iris, Species)
```

```{r}
#To select several variables with tidyverse
dplyr::select(iris,Species, Sepal.Length, Sepal.Width)

# To select several lines inside data frame
dplyr::slice(iris, 22:30)
# I can affect it to an object

subdata <- dplyr::slice(iris, 22:30)

#or choosing different lines
myline <- c(22, 38, 120) # I build a vector containing the line that I want
dplyr::slice(iris,myline)
```
##### Bonus : to pipe many function serval function together
The pipe <span style="color: steelblue;"> **|>** </span>, or <span style="color: steelblue;">**%>%**</span>, is one of the useful elements of the tidyverse. It allows to structure sequences of operations by minimizing the creation of intermediate objects and by facilitating the addition of a step anywhere in this sequence.

The useful command to manage data frame : 1) select() to select vector or variable from a data frame ; 2) filter() is used to subset a data frame, retaining all rows that satisfy your conditions ; mutate() adds new variables and preserves existing ones.

```{r}
library(dplyr)

iris %>%
  filter(Sepal.Length > 6) %>% # filtered using size of Sepal Length
  filter(Species == "versicolor") # and the species

```

```{r}
iris %>%
  select(Sepal.Length, Species) %>%
  mutate(Sepal.Length2 = Sepal.Length * 2) %>%
  mutate(Sepal.Length2_squared = Sepal.Length2 * Sepal.Length2)

iris %>%
  select(Sepal.Length, Species) %>%
  mutate(Sepal.Length = Sepal.Length / mean(Sepal.Length, na.rm = TRUE))

#Sepal.Length_norm is calculated using the mean of the Sepal.Length for all data set.

iris %>%
  select(Sepal.Length, Species) %>%
  group_by(Species) %>%
  mutate(Sepal.Length_norm = Sepal.Length / mean(Sepal.Length, na.rm = TRUE))

# In this last example Sepal.Length_norm is calculated using the mean of the Sepal.Length for each species using group_by()
```

**<span style="color: red;"> These commands are non-persistent, no changes are made on the original iris data frame.</span>**
**<span style="color: red;">If you want to store it, you must assign your changes to a object</span>**

```{r}
iris_modif <- iris %>%
  select(Sepal.Length, Species) %>%
  group_by(Species) %>%
  mutate(Sepal.Length_norm = Sepal.Length / mean(Sepal.Length, na.rm = TRUE))
```

#### How to import external data frame issue from .txt or .csv

To import data set, the function `read.table()` or `read.csv()` are commonly used.

`read.table(file, header = FALSE, sep = "", dec = ".", ...)`

The main parameter are :

- **file** : add the pathway and the name of the file
- **header** : a logical value (TRUE or FALSE) indicating whether the file contains the names of the variables as its first line. 
- **sep** : the field separator character. Values on each line of the file are separated by this character. If sep = "" (the default for read.table) the separator is ‘white space’, that is one or more spaces, tabs, newlines or carriage returns.
- **dec** : the character used in the file for decimal points.

```{r}
ds <- read.table(here::here("data",
                            "rforbeginers",
                            "exemple_read.txt"),
                 header = TRUE,
                 sep = ";",
                 dec = ",")
```


<span style="color: steelblue;">Why I use the parameter `header=TRUE` ?</span>

#### To export data set as .txt to read in excel

The function is similar to `read.table()`

`write.table(x, file = "", sep = " ",na = "NA", dec = ".", ... )`

- **x** : this is your data.frame
- **file** : give a name for your file
- **sep** : cf read.table
- **dec** : cf read.table
- **na** : give a symbole for missing data, by convention is NA

```{r}
write.table(ds, "ds.txt", sep = "\t", dec = ".")
```

To keep our working directory tidy, we now delete `ds.txt`

```{r}
file.remove("ds.txt")
```
Exo 4

1. In the dataset Iris select Sepal Width, Sepal length and  Species,to create a new data frame name "subset_iris"
2. Save this new data frame as text file 

## How to represent your data

**ggplot2** is a powerfull packages to make a very smart graph "ready to use" for publication. **gg** means grammar and graph, a concept which describe a graph using grammar. This package belong to tidyverse according to `dplyr`. According to the ggplot2 concept, a graph can be divided into different basic parts:**Plot = data + Aesthetics + Geometry**

- *data* : data frame
- *aesthetics* : allows to indicate the x and y variables. It can also be used to control the color, size and shape of the points, etc...
- *geometry* : corresponds to the type of graph (histogram, box plot, line plot, .....)

```{r}
library(ggplot2)
# scatter plot 
ggplot(data = iris, aes(Sepal.Length, Sepal.Width)) + #  what I want to represent 
  geom_point() # how I represent it
```

```{r}
# We can calorize and give a shape by month
# scatter plot
ggplot(data = iris, aes(Sepal.Length, Sepal.Width)) +
  geom_point(aes(color = Species, shape = Species)) # shape is the same thing with classical plot on R
```
```{r}
# You can change manually color with la fonction scale_color_manual()
ggplot(data = iris, aes(Sepal.Length, Sepal.Width)) +
  geom_point(aes(color = Species, shape = Species)) +
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))
```
```{r}
#Boxplot
ggplot(data = iris, aes(Species, Sepal.Length)) + #  what I want to represent 
  geom_boxplot(aes(color = Species, fill = Species), alpha=0.4) + # how I represent it. Here I changed geometry using boxplot.
  labs(x ="Species", y = "Sepal.Length (mm)") +
  theme_minimal()
```

**Save your plot**

Using the function `pdf()`

Firstly you have to store plot inside and object

```{r}
#Boxplot
g1 <- ggplot(data = iris, aes(Species, Sepal.Length)) + #  what I want to represent 
  geom_boxplot(aes(color = Species, fill = Species), alpha = 0.4) +# how I represent it. Here I changed geometry using boxplot.
  labs(x = "Species", y = "Sepal.Length (mm)") +
  theme_minimal()
```

```{r}
pdf("yourfile.pdf")
g1
dev.off()
```

or `ggplot2::ggsave()`

```{r}
ggsave("yourfile2.pdf", g1)
```
