---
title: "Part 0: R for beginner"
output: html_document
author: "Marc Garel, MIO"
editor_options: 
  chunk_output_type: inline
---

## How to install R?
### Useful link

To install Cran R, you must go to [https://cran.r-project.org/](https://cran.r-project.org/), 
and download the install file for your favorite operating system,
click on the .exe, .dmg, .pkg, .deb, respectively for Windows, MacOS and Linux-debian.

And click on follow... until to reach successful installation
Then download and install the IDE Rstudio
[https://www.rstudio.com/products/rstudio/](https://www.rstudio.com/products/rstudio/).

*Everything is free to download*

### First commands

**Where am I?**

To get the current working directory

```{r}
getwd()
```

**To change my working directory from the console**

```{r eval=FALSE}
setwd("/path/to/my/fancy/project/")
```

In Rstudio, we can change the working directory by navigating in folder in **File** panel
and clicking on menu **More**

If you are using a Rproject, you don't need to change your working directory.

**How can I find documentation about function?**

The command `help()` is the 911

```{r eval=FALSE}
# exemple with function read.table
help("read.table")
#other exemple of 911
example("plot")
??plot()
```


## How to use a package?
### Installation

You can install new packages by clicking directly in Rstudio
or by command line (the best way for me)

```{r eval=FALSE}
install.packages("your_package") # for packages on CRAN mirror
```

For packages from Bioconductor (specifically for bioinformatic): 

```{r eval=FALSE}
BiocManager::install("your_package")
```

or from github (using the package devtools)

```{r eval=FALSE}
devtools::install_github("your_package")
```

### Using functions from a package

To use functions from a specific package you can either load the entire package:

```{r eval=FALSE}
library("your_package")
```

or call the function this way:

```{r eval=FALSE}
your_package::yourfunction()
```


## Expression and affectation
### Expression

An expression is directly evaluated and the result is displayed on terminal
Example : 

```{r}
2 + 3
sqrt(25)
```


### Affectation in an object
An assignment is an expression stored in object or variable.
In this example expression, constant, array, matrix, data frame, list
Example :

```{r}
a <- 2 + 3
```

You can also assigned a expression to variable using `=`, like in most
of other languages.

```{r}
b = 3 + 3
```

There are debates about which one to use.
It is not really important which one you use, just be
consistant and always use the same way to assign.

To list object/variable from your environment:

```{r}
ls()
```

### Type of objects

#### Vectors

Vectors are objects composed by values with the same type (i.e, numeric, characters...)

```{r}
V1 <- c(2, 6, 9) # numeric vector
V2 <- c("monday", "Tuesday", "Wednesday")# character vector
V3 <- rep(6, 3) # repetition of the same value
V4 <- seq(1, 3, 0.1)
V5 <- 1:100
V5
```


To know the number of value inside a vector

```{r}
length(V1)
```

**What do you think about V4?**

Filter a vector according to criteria

```{r}

# Example : with list of value (a vector)

x <- c(1, 3, 5, 3, 2, 1, 4, 6, 4, 7, 5, 4, 3)

# get element from 2 to 6

x[2:6]

# get elements 3 et 5 from x.

x[c(3, 5)]

# get value more than 20.

x[x > 20]

# get value of  x where x is equal to 21.

x[x == 21]

# return elements form x the the value different from 5

x[x != 5]
```

Filter a vector according to several criteria

```{r}
# 3 lists : ages, sexes et poids

age <- c(20, 30, 40,
         15, 22, 24,
         36, 38)

sexe <- c("F", "M", "F",
          "M", "F", "M",
          "F", "M")

poids <- c(75, 76, 73,
           72, 64, 76,
           73, 72)

# get value from age greater than 20 and less than 30.

age[age > 20 & age < 30]

# Recovering "poids" for those who are older than 25 and female

poids[age > 25 & sexe == "F"]

#Retrieve age values below 20 or above 30.

age[age < 20 | age > 30]
```

*Exo1*

Considering the vector `a` such as 
`a <- c("lannister", "targaryen", "baratheon", "starck", "greyjoy")`

1. What is the length of the vector?

2. Try doing `a[1:3]`. What do you get?

3. Create a new vector b containing only lannister and starck.

4. Try doing `a[-1]`. What do you get?

5. Sort by alphabetical order using `sort()`

*Exo2*

1. Create a vector `a` containing all integers from 1 to 100.

2. Add the values 200, 201, 202 to the vector `a`.

3. Create a vector `b` containing all even integers from 2 to 100 using `seq()`

#### Matrices

Matrices are objects in which the elements are arranged in a two-dimensional rectangular layout.
As for vector all the elements are of the same type.

```{r}
m1 <- matrix(1:12, ncol = 3, byrow = TRUE)
m2 <- matrix(1:12, ncol = 3, byrow = FALSE)
m1
# to know the dimension of the matrix (number of line and number of column)
dim(m1)
ncol(m1)
nrow(m1)
```

**Some operation on matrix**

```{r}
#Selection all data inside 2 column
c2 <- m1[, 2]
c2

#Select data inside 3rd column and 4th line
m1[4, 3]

#delete a column
m3 <- m1[, -2]
m3
m3[2]
```

*Exo3*

1. Run the command `a <- rep(c(0,1), 50)`. What happened?

2. Use `a` to construct a matrix `A` with 10 rows and 10 columns. 

3. Print the dimensions of this matrix.

4. Use `t()` on this matrix to create a matrix `B`. What happened?

5. The commands `A[1:5, ]` and `B[, 1:5]` retrieve the first 5 rows of A 
and the first 5 columns of B. Use these commands to retrieve the rows containing
only 1s from A and the columns containing only 0s from B.

*Exo 4 sup*

1. Create matrix M

$$M = \begin{bmatrix}3 & 5 & 6\\
-1 & 2 & 2 \\
1 & -1 & -1 \\
\end{bmatrix}$$

2. Display the dimension of `A`, its number of columns, its number of rows and its length
3. Extract the second column of `A`, then the first row
4. Extract the element at the third position of the first line;
5. Compute the sum of the columns and rows of `A` using `colSums()`, `rowSums()`

#### Data frames

Data frames are objects composed by vector where the value
are of different modes (i.e, numeric, characters...)

##### Data frame examples

Load a data frame

```{r}
data(iris)
```

Visualise the data frame in a table

```{r eval=FALSE}
View(iris)
```

Display its internal structure

```{r}
str(iris)
```

**What can we notice?** 

##### How to build your own data frame

```{r}
date <- c("1_monday", "2_Tuesday", "3_Wednesday",
          "4_Thursday", "5_Friday", "6_Sturday",
          "7_Sunday")

is.character(date)

# temperature in deg Celsius
temperature <- c(24, 27, 25,
               22, 30, 21,
               28)

is.numeric(temperature)

# rain in mm
rain <- c(1, 0, 0,
          5, 2, 0,
          0)

is.numeric(rain)

# add column together
df <- cbind(date, temperature, rain)
str(df)

# make data.frame
df <- data.frame(date, temperature)
str(df)

df$rain <- rain

#Give me the difference between cbind() and data.frame()
View(df)

#To select a column or vector
df$temperature
df[, 2]

# here we use list() instead of c()
# because there is multiple class in inside row

day <- list("8_monday", 29, 1)
new_def <- rbind(df, day)# add row to a data frame
new_def
```

##### Use dplyr to select, filter a data frame

`dplyr` is part of the library set named `tidyverse`
(contraction of "tidy" and "universe", it's a tidy universe). 
`tidyverse` packages are designed to work together and thus follow
the same code logic and a common grammar.

The pipe, `%>%`, is one of the useful elements of the tidyverse. 
It allows to structure sequences of operations by minimizing 
the creation of intermediate objects and by facilitating 
the addition of a step anywhere in this sequence. Note that from
R 4.1 you can use a new pipe, `|>` without the need of loading any
library.

The most commonly used tidyverse packages are loaded in your session:

- `ggplot2`
- `dplyr`
- `tidyr`
- `readr`
- `tibble`
- `stringr`


```{r}
tidyverse::tidyverse_packages()
```


```{r}
library(tidyverse)

#To select a variable with tidyverse
#Command line widely used to select data
iris$Sepal.Width %>%
  head()
iris[, "Sepal.Width"] %>%
  head()

#With dplyr
select(iris, Sepal.Width) %>%
  head()

```

```{r}
#Command line widely used to select data
iris[, c("Sepal.Length", "Sepal.Width", "Species")] %>%
  head()
#With tidyverse command
iris %>%
  select(Sepal.Length, Sepal.Width, Species) %>%
  head()
```

To sort and select data inside data frame

```{r}
# Command line widely used to select data
iris[which(iris$Sepal.Width > 3 & iris$Species == "virginica"), ] %>%
  head()

#With tidyverse command
iris %>%
  filter(Sepal.Width > 3) %>%
  filter(Species == "virginica") %>%
  head()
```


## Functions
### Definition
Function are a compilation of command line with different instructions 
inside one object to simplify code. A function is composed by
 <span style="color: steelblue;">**arguments**</span> and <span style="color: steelblue;">**options**</span>.

function(argument1, argument2, option1, ... ,option10)


### Usual functions for data frame

```{r, eval=FALSE}
head() # to know first line of your data frame
class()# return the class of the object. Ex : data.frame, matrix, list ....
str()# return the structure of the object. Ex : numeric, factor, character....
names()# to get or set the names of an object
sum() # for addition
min() # return the minimum of the vector
max() # return the minimum of the vector
row.names() # attribute names for lines of the data frame
colnames() # attribute names for column of the data frame
apply() # Returns a vector or array or list of values obtained by applying a function to margins of an array or matrix.
```

**Example function row.names, class and str**

```{r}
df2 <- data.frame(x = c(TRUE, FALSE, NA, NA), y = c(12, 34, 56, 78))
df2
row.names(df2) <- paste("row", 1 : 4, sep = "_")
df2 # what do you see
class(df2)
str(df2)
```

**Example function apply**

```{r}
head(df)
class(df)
str(df)
# return mean for the numerical column of the data.frame. apply(data,margin,fun). 
# For margin parameter the value 1 return mean for each row, 
# for margin=2 return mean for each selected column.
apply(df[, 2:3], 2, mean)
```

**Example function plot**

The function `plot()` is a very useful function to make a graph for examination for your data set

Let's have a look at the documentation of this function `?plot`

*Exo 5*

Load data set iris using `data()`

1. Give the class Sepal.Width and Species vectors
2. What is the minimum / maximum / average sepal length of these irises?
3. What is the values of the 10 first sepals?
4. Calculate standard deviation for every numeric vector
5. Calculate standard mean for every numeric vector
6. Create a data frame with mean and sd as line and give a name for each line
7. An error of 0.5cm was made in the measurement of the length of the sepal of the 41st iris: add 0.5cm to this measurement
8. Plot Sepal.Length as fonction of Sepal.Width adding title "Sepal" xlab "Sepal.Length" and ylab "Sepal.Length"
9. Run once again plot giving the following parameters pch=19, cex=2, col="red", cex.axis=2, cex.lab=2. What do you see?


#### How to import external data frame issue from .txt or .csv

To import data set, the function `read.table()` or `read.csv()` are commonly used.

`read.table(file, header = FALSE, sep = "", dec = ".", ...)`

The main parameter are :

- **file** : add the pathway and the name of the file
- **header** : a logical value (TRUE or FALSE)indicating whether the file contains the names of the variables as its first line. 
- **sep** : the field separator character. Values on each line of the file are separated by this character. If sep = "" (the default for read.table) the separator is ‘white space’, that is one or more spaces, tabs, newlines or carriage returns.
- **dec** : the character used in the file for decimal points.

```{r}
ds <- read.table(here::here("data",
                            "rforbeginers",
                            "exemple_read.txt"),
                 header = TRUE,
                 sep = ";",
                 dec = ",")
```


<span style="color: steelblue;">Why I use the parameter `header=TRUE` ?</span>

#### To export data set as .txt to read in excel

The function is similar to `read.table()`

`write.table(x, file = "", sep = " ",na = "NA", dec = ".", ... )`

- **x** : this is your data.frame
- **file** : give a name for your file
- **sep** : cf read.table
- **dec** : cf read.table
- **na** : give a symbole for missing data, by convention is NA

```{r}
write.table(ds, "ds.txt", sep = "\t", dec = ".")
```

To keep our working directory tidy, we now delete `ds.txt`

```{r}
file.remove("ds.txt")
```

## The world of "PLOT"

We load data set included in R 

```{r}
data("airquality")
```

How can I check data ? before to plot, could you check the structure of data.frame

```{r}
plot(airquality$Month, airquality$Ozone)
```

add label for each axes, change the symbol by square and color blue

```{r}
boxplot(airquality$Ozone ~ airquality$Month,
        col=c("orange", "yellow", "blue", "red", "green"))
```

**Use of ggplot2 packages to generate smart plot**

`ggplot2` is part of tidyverse, so it is already loaded

```{r}
g1 <- ggplot(data = airquality) +
  geom_point(aes(Month, Ozone))

g1

g2 <- ggplot(data = airquality) +
  geom_point(aes(Month, Ozone), size = 3, colour = "red") +
  theme_bw()

g2
```

What do you notice? 

```{r}
df <- data.frame(airquality$Month, airquality$Temp)
df$airquality.Month <- as.factor(df$airquality.Month)
g3 <- ggplot(data = df, aes(x = airquality.Month, y = airquality.Temp)) +
  geom_boxplot() +
  theme_bw()
g3

g4 <- ggplot(data = df, aes(x = airquality.Month,
                            y = airquality.Temp,
                            color = airquality.Month)) +
  geom_boxplot() +
  theme_bw()
g4

```

**Save your plot**

Using the function `pdf()`

```{r}
pdf("yourfile.pdf")
g4
dev.off()
```

or `ggplot2::ggsave()`

```{r}
ggsave("yourfile2.pdf", g4)
```

```{r}
list.files(pattern = ".pdf") %>%
  file.remove()
```

```{r}
data("BCI", package = "vegan")
chart <- apply(BCI, 2, sum)
selection1 <- chart[1:30]
dotchart(selection1, cex = 0.7, bg = "blue")
row.label <- names(selection1)
df2 <- data.frame(row.label, selection1)

ggplot(df2) +
  geom_point(aes(x = selection1, y = row.label))


g6 <- ggplot(df2, aes(x = selection1, y = row.label)) +
  labs(tag = "A", x = "Species", y = "Count")

g6 +
  geom_bar(stat = "identity")

g6 +
  geom_bar(stat = "identity", color = "lightblue", bg = "lightblue")
```

Save this plot using command `ggsave()`. The documentation is
your best friend. 

*Exo6*

We propose to evaluate the effect of a vitamin treatment supposed
to make students more resistant to fatigue during exams.

To do this, different students are tested: a score out of 10
is attributed to them and expresses the degree of rest called
here "Acuity", 0 for very tired and 10 for in full form.

10 students take a treatment (Treatment = "O" for "Yes")
and their fatigue is evaluated every day during 5 days
of intense revision where the fatigue increases progressively.

10 students take a placebo (Treatment = "N" for "No") and
their fatigue is assessed every day for 5 days.

1. Load the table "exercise6.txt". This table has a header and a tab separation.

2. Plot the tiredness (here "Fatigue") as a function of time for all tested individuals.
Display the treated individuals in blue and the untreated ones in red (see example).
(*To filter data set use `by()` function*)

3. Try with ggplot

4. Save your plot as a pdf